---
title: "R Notebook"
output: 
    html_notebook:
        toc: true
        toc_depth: 3
---

# The very basics

## Getting Help

At the command line we can look for help through the commands:

-   `help()`: gets information on a particular R function, e.g. `help(matrix)`

-   `?`: works the same as the previous one, e.g. `?"matrix"`

-   `help.search()`: It's useful if we know the subject rather than the name.

-   `apropos()`: returns a vector of names of objects containing that string, e.g. `apropos("matrix")` returns every object with 'matrix' in its name.

-   `args()`: displays the set of arguments expected by a given function.

```{r}
args(matrix)
```

# R Data: Vectors

A vector is the basic unit of computation in R. The atomic clases which we can see with `help(vector)` are::

-   logical: TRUE or FALSE (T or F as synonyms)
-   integer
-   numeric or double
-   character: is the text
-   raw: can hold binary data
-   complex

To combines items into a vector we use `c()` function. Cal also be used to combine vector, as long as all the vectors are of the same sort.

```{r}
c(2, 25)
```

```{r}
# R has formatted the numbers in a consistent way
c(-2, pi, 25)
```

```{r}
c(c(2, 3), c(pi, 16.7, 5), c(-12, -16))
```

To set consecutive integers:

```{r}
1:7
```

```{r}
# Sequence in reverse
9:-2
```

```{r}
# It's 3 plus the sequence
3 + 2:5
```

Using `help(Syntax)` It can be seen a list of operators and their precedences, and precedence can always be over-ridden with parentheses.

Also with `seq()` we can create sequences, with better control of starting points, ending points, lengths, and step sizes. With help we can see:

``` r
seq(...)

## Default S3 method:
seq(from = 1, to = 1, by = ((to - from)/(length.out - 1)),
    length.out = NULL, along.with = NULL, ...)

# Examples:
seq(17)
seq(0, 1, length.out = 11)
seq(1, 9, by = 2)
```

Logical vectors ofen are constructed by R in response to an operation:

```{r}
35:41 >= 37
```

To calculate the number or proportions of elements in a vector we use `sum()` for count and `mean()` for proportions:

```{r}
# like we saw, five out of seven elements in this vector are greater or equal to 37
sum(35:41 >= 37)
```

```{r}
# 5 out of 7 are aprox. 71.4%
mean(35:41 >= 37)
```

## Vector Operations

Arithmetic operations on vectors produce vectors.

```{r}
5:10
```

```{r}
(5:10)^2
```

There are functions which returns a single number (that is also a vector), such as: `length(), sum(), mean(), sd(), min(), max()`. The function `range()` returns the smallest and largest values, and `summary()` returns a vector of summary statistics.

```{r}
# An assigment with parentheses case print this assigment
(th1 <- c(20, 15, 10, 5, 0)^2)
```

```{r}
th2 <- 105:101
th2 + th1
```

```{r}
th2 / th1
```

It's Computed element by element, that's why print 'Inf', because 101/0 refer to an infinite value.

```{r}
# Print the larger value anywhere
max(th1, th2)
```

```{r}
# 'Parallel maximum' print the larger of each pair
pmax(th2, th1)
```

Loginac vectors can also be combined. The operator `|` for 'or' return TRUE if either element is TRUE; the `&` operator for 'and' return TRUE only if both elements are TRUE. This single version evaluate the condition for every pair of elements from both vectors, the double version ( `|| , &&`) evaluate multiple TRUE/FALSE conditions from left to right, stopping as soon as possible (are useful in `if()` statements).

Recycling

When a vector does not match with other vector to be operated, the shorter one starts again, recycling it's elements.

```{r}
# Two vector with the same length, 6
5:10 + c(0, 10, 100, 1000, 10000, 100000)
# The sum is element by element
```

```{r}
# Two vectors, 6 and 3 elements
5:10 + c(1, 10, 100)
```

```{r}
# 6-vector and 5 vector. The second recycle it's first element (10 + 3)
5:10 + c(3, 4, 5, 6, 7)

```

Recycling a vector of length 1 (`(5:10) + 4`) is very common, but the other lengths ir rarer. If we see the last warning, It's better to treat that as an error and try to resolve it.

## Handling character vectors
