---
title: "Data Science for Business"
author: "Alvaro Manzanas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
url: "https://cognitiveclass.ai/learn/data-science-business"
output:
  github_document:
    md_extensions: +autolink_bare_uris+tex_math_single_backslash
    fig_width: 12
    fig_height: 8
    df_print: kable
    number_sections: true
    toc: true
    toc_depth: 3
  rmdformats::html_clean:
    code_folding: hide
    self_contained: true
    number_sections: true
    thumbnails: false
    lightbox: true
    highlight: breezedark
subtitle: "Analysis Procedure"
csl: apa.csl
pkgdown:
  as_is: true
editor_options:
  markdown:
    wrap: 79
---

# IBM ML0103EN. Digital Analytics & Regression

## Introduction

-   What is the goal of the solution?
-   Why do we need to do this?
-   What data is available?
-   What constraints exist?
-   What is an acceptable solution?
-   What is success in this context?
-   How do we measure success?

Business objectives drive data science initiatives. 

> When solving business problem with data, be curious!

1.  Objective, problem, decision. Think of the business problemas as a statement to accept or reject an hypothesis. e.g., *The product name should be option A because GoogleTrends showed more data searches*. That will be the hypothesis to work.

2.  Context. Working in a new project with a team, take in as much information from the team as possible. But careful, accept nothing as fact that you cannot reaonably independently verify, because team members information may come with biases.

3.  What are the options. Mentionen in the Hypothesis, we can Option A, and Option B.

4.  What have we not yet considered. Are there any other options that the initials two? There are more data in GoogleTrends which we can analyse? Which option relects more accurately the audience.

5.  What are the constraints. Like Budget, time, reliability of the data, legal and ethical constraints...

6.  What data de we have. If already we have data to work with.

7.  What data de we need to get. Extract, mine or collect data to this purpose.

## Importing GoogleTrends data in R

In GoogleTrens look for a topic of interest https://trends.google.com/trends/ and download the CSV.

```{r}
df <- read.csv("../data/report.csv", header = F)
head(df)
```

## Simple Regression in R

```{r}
###########################################################################
#                    Upload & Transform a Google Analytics .csv file
###########################################################################

case_data <- "../data/report_2_.csv"
open_file  <- file(case_data, open = "r")

linecount <- 0
string_data <- ""
while (length(single_line <- readLines(open_file, n = 1, warn = FALSE)) > 0) {
  linecount <- linecount + 1
  
  if (linecount < 3) {
    case_data <- paste0(case_data,single_line)     
  }
  
#Typical Google Analytics CSV outputs have the column headings in row 5
  if (linecount == 5) column_headings = strsplit(single_line, ",")[[1]]
  
#We do not need the first 5 lines in the CSV
  if (linecount > 5) {

#Typical Google Analytics CSV outputs have a blank row after the first data set
    if (gsub(pattern=",", x=single_line, replacement="") == "") break
    
    string_data <- paste0(string_data,single_line,"\n")
  }
}
close(open_file)

case_table <- read.table(textConnection(string_data), sep=",", header=FALSE, 
                         stringsAsFactors = FALSE)

#Assigning the column headings to our data table
names(case_table) <- column_headings

```

```{r}
###########################################################################
#                    Let's look at the data
###########################################################################

head(case_table)
column_headings

#Instead of the complex 'Week' column, let's add an ID column
case_table$WeekID<-seq.int(nrow(case_table))

head(case_table)

#We can also look at the data using
#print(case_table)

#Or if you're used to spreadsheet formats, this output looks like a table
#NB: The 'table' appears in a different tab in the tool bar so you would need to come back to this tab after viewing your data
#View(case_table)

#Or
#case_table

#With big data, you would not want to call all the rows in your data set
#so you can call a specific number of lines
head(case_table, n = 10L)
#Brings back the first 10 lines, n = 10L
#You can also only bring back the summary for one column using the $ sign
head(case_table$WeekID, n = 10L)

#We can now start plotting the different trends!
```

```{r}
##############################################################
#                    Plotting Google Trends in R
##############################################################

plot(case_table$WeekID, case_table$analytics, type='l', col='orange')
lines(case_table$WeekID, case_table$`google analytics`, type='l', col='red')
```

```{r}
###########################################################################
#   Which search term has the highest search volumes in Google Trends?
###########################################################################

par(mfrow = c(1, 1))
plot(case_table$WeekID, case_table$analytics, 
     main = "Searches for 'Analytics' & 'Data Scientist'",
     ylab = "Search Index (0 to 100)",
     xlab = "Week Index",
     type = 'l',
     col  = 'orange')
lines(case_table$WeekID, case_table$`data scientist`, type = 'l', col = 'red')
```

```{r}
#Is 'data scientist' too narrow? It has very few searches, what about 'data science'?
par(mfrow = c(1, 1))
plot(case_table$WeekID, case_table$analytics, 
     main = "Searches for 'Analytics' & 'Data Science'",
     ylab = "Search Index (0 to 100)",
     xlab = "Week Index",
     type = 'l',
     col  = 'blue')
lines(case_table$WeekID, case_table$`data science`, type = 'l', col = 'darkgreen')
#See the new plot on in the bottom right hand quadrant
```

```{r}
#Is there a difference between 'data scientist' and 'data science'?
par(mfrow = c(1, 1))
plot(case_table$WeekID, case_table$`data scientist`, 
     main = "Searches for 'Data Scientist' & 'Data Science'",
     ylab = "Search Index (0 to 100)",
     xlab = "Week Index",
     type = 'l',
     col  = 'red')
lines(case_table$WeekID, case_table$`data science`, type = 'l', col = 'blue')

#All versions of 'data science' are exclipsed by searches for 'analytics'
#The Search Index (0 to 100) only goes up to 4 out of a maximum value of 100 for 'data scientist' and 'data science'

```

```{r}
###########################################################################
#                    Let's look at correlation
###########################################################################

#R's default correlation is 'pearson' if none is specified
#?cor

#What is the relationship between searches for 'data science' and 'data scientist'?
#Can we quantify that relationship?
cor(case_table$`data science`,case_table$`data scientist`)
#The answer will appear in the bottom left quadrant in the RStudio Console below

#Round off decimal points
round(cor(case_table$`data science`,case_table$`data scientist`), digits = 2)
#The correlation for data scientist and data science terms is 88% so we will conclude that 
#People are using these search terms in a similar way, in a similar pattern, in similar weeks

#What is the correlation between 'analytics' and 'machine learning'
round(cor(case_table$`analytics`,case_table$`machine learning`), digits = 2)
#What is the correlation between 'data science' and 'machine learning'

round(cor(case_table$`data science`,case_table$`machine learning`), digits = 2)
#There is a strong positive correlation between 'data science' and 'machine learning'
#There is a weak correlation between 'analytics' and 'machine learning'

#Let's work with the two variables, 'data science' and 'analytics' 
par(mfrow = c(1, 1))
plot(case_table$WeekID, case_table$analytics, 
     main = "Searches for 'Analytics' & 'Data Science'",
     ylab = "Search Index (0 to 100)",
     xlab = "Week Index",
     type = 'l',
     col  = 'blue')
lines(case_table$WeekID, case_table$`data science`, type = 'l', col = 'red')


#Can we quantify this relationship?
round(cor(case_table$analytics,case_table$`data science`), digits = 2)
#The correlation is only 18%, we can conclude that these two searches are not happening in a similar way
#These may be different people searching or people searching for different content

```

```{r}
###########################################################################
#                    Summary statistics
###########################################################################

#Summary statistics for each variable
summary(case_table)

#Let's make it easier to work with the two search terms we are interested in 'analytics' and 'data science'
#We'll shorten how we refer to them by creating new variables
Analytics <- c(case_table$analytics)
DataScience <- c(case_table$`data science`)

#To get more statistical information like skew and kurtosis, we use a package called 'psych'
install.packages("psych")
#This may take a minute
library(psych)

describe(Analytics)
describe(DataScience)

#The search behavior for the two terms is very different. 
#How can we learn more about the individual distributions?

```

```{r}
###########################################################################
#                    Box plots
###########################################################################

par(mfrow = c(1, 1))
boxplot(Analytics, DataScience, 
        names = c("Analytics","DataScience"),
        xlab  = "Search Terms",
        ylab  = "Search Index (0 to 100)",
        main  = "Boxplot of search terms",
        col   = "orange")

```

```{r}
###########################################################################
#                    Histograms
###########################################################################

par(mfrow = c(2, 1))
hist(Analytics,
     main = "Histogram of Searches for 'Analytics'",
     ylab = "Frequency of searches",
     xlab = "Search Index (0 to 100)",
     col  = "orange")
hist(DataScience,
     main = "Histogram of Searches for 'DataScience'",
     ylab = "Frequency of searches",
     xlab = "Search Index (0 to 100)",
     col  = "blue")
#Search Index Number is between 1 and 100

```

```{r fig.height=12, fig.width=8}
###########################################################################
#                    Scatter plots
###########################################################################

par(mfrow = c(2, 1))
plot(Analytics,
     main = "Scatter plot of searches for Analytics",
     col  = "maroon")
plot(DataScience,
     main = "Scatter plot of searches for Data Science",
     col  = "darkgreen")
plot(Analytics,DataScience,
     main = "Scatter plot of searches, Analytics vs. DataScience",
     col  = "orange")

#What does that look like? We compare 'data science' and 'machine learning'
plot(DataScience, case_table$`machine learning`,
     main = "Scatter plot of searches, 'DataScience' vs. 'Machine Learning'",
     col  = "darkblue")
#The few data points in the plot look positive correlated, and we know that they are
round(cor(case_table$`data science`,case_table$`machine learning`), digits = 2)
#There is a strong positive correlation between 'data science' and 'machine learning'
# 90%

```

```{r}
###########################################################################
#                    Strong positive correlation
###########################################################################

#Of the Google Trends data we pulled, is there anything else in there that is highly correlated to 'analytics'?

#Remember this?
#What data is actually being pulled?
head(case_table)

#Take a closer look at the other columns in the extraction
column_headings

#Let's make it easier to work with the term 'Google Analytics'
#We'll shorten how we refer to it by creating a new variable
GoogleAnalytics <- c(case_table$`google analytics`)

#Let's test whether people searching for 'analytics' are likely searching for 'google analytics' related content
par(mfrow = c(1, 1))
plot(case_table$WeekID, case_table$analytics, 
     main = "Searches for 'Analytics' & 'Google Analytics'",
     ylab = "Search Index (0 to 100)",
     xlab = "Week Index",
     type = 'l',
     col  = 'blue')
lines(case_table$WeekID, GoogleAnalytics, type = 'l', col = 'red')

#Can we quantify this relationship?
round(cor(Analytics,GoogleAnalytics), digits = 3)

#A perfect correlation is 100%, these two terms have a near perfect correlation!
#These searches are consistent with searches more about digital marketing, search engine marketing and general analytics

#Let's see if the distributions for the two terms are similar?
#Recall the package 'Psych'?

describe(Analytics)
describe(GoogleAnalytics)

#The search behavior for the two terms is very similar. 
#How can we learn more about the individual distributions?

```

```{r}
###########################################################################
#                    Visualizing the strong positive correlation
###########################################################################

par(mfrow = c(1, 1))
boxplot(Analytics, GoogleAnalytics, 
        names = c("Analytics","Google Analytics"),
        xlab  = "Search Terms",
        ylab  = "Search Index (0 to 100)",
        main  = "Boxplot of search terms",
        col   = "maroon")

#**********Look at histograms**********
par(mfrow = c(2, 1))
hist(Analytics,
     main = "Histogram of Searches for 'Analytics'",
     ylab = "Frequency of searches",
     xlab = "Search Index (0 to 100)",
     col  = "orange")
hist(GoogleAnalytics,
     main = "Histogram of Searches for 'Google Analytics'",
     ylab = "Frequency of searches",
     xlab = "Search Index (0 to 100)",
     col  = "grey")


#**********Look at scatter plots**********
par(mfrow = c(1, 1))
plot(GoogleAnalytics, Analytics,
     main = "Scatter plot of searches, Analytics vs. Google Analytics",
     col  = "orange")
#Note that 'Analytics' is on the 'y' axis, that is known as the 'dependent variable'

#Add a line of best fit
abline(lm(formula = Analytics ~ GoogleAnalytics), main = "Scatter plot of searches, Analytics vs. Google Analytics", col="red") 


```

```{r fig.height=12, fig.width=12}
###########################################################################
#                    Regression
###########################################################################

#To fit an ordinary linear model, we use the 'lm' function for 'linear model'
#We assign the results to an object called 'Regression'
Regression <- lm(formula = Analytics ~ GoogleAnalytics)
#The first variable is the target variable and the second is the predictor variable or explanatory variable
#We are looking to see if variations in the variable 'Analytics' can be explained by variations in the variable 'GoogleAnalytics'


#Let's look at it in the  console below: 
Regression

#Round output to 2 decimal places
#Round(x,digits = 2)

round(coef(Regression), digits = 2)

#Regression output
summary(Regression)
round(summary(Regression)$coefficients, digits = 3)
round(summary(Regression)$r.squared, digits = 3)

par(mfrow = c(1, 1))
plot(GoogleAnalytics, Analytics,
     main = "Scatter plot of searches, Analytics vs. Google Analytics",
     col  = "orange")

plot(Regression)
#Show all 4 charts in the same view using 'par'
par(mfrow = c(2,2))

#Now for the Regression plots:
plot(Regression)

abline(Regression)

#Note the 'Zoom' button in the 'Plots' tab in the bottom right hand quadrant 
#'Zoom' will open the chart in a new browser window, you may get be asked to allow the 'pop-up'

#We can call out specific values of interest from the 'Regression' output e.g., r squared
round(summary(Regression)$r.squared, digits = 2)
#Note the near perfect R-Squared in the output
#Note the statistically significant P-Value

#What else can you pull out?
names(summary(Regression))

#So what is our model or line of best fit formula?
Regression
#Y = 4.132 + 1.751X

#This formula can be read as: analytics = 4.140 + 1.754googleanalytics 
#What search index number for 'analytics' can we expect when the searche index for 'googleanalytics' is 33?
4.132 + (1.751*33) 

#The answer will appear in the bottom left quadrant in RStudio

#Round it off to one decimal point.
round(4.132 + (1.751*33), digits = 1)


```

## Presenting Data Analytics in Business

This is the final step, to communicate our conclusions and recommendations effectively to a professional business audience.





























